%% ===================================================================
\section{Introduction}
\label{sec:intro}
%% ===================================================================

% Par 1: SP motivation, gap
Closed-form models of nonlinear signal-processing systems are
useful for interpretability, sensitivity analysis, and downstream
design.  Symbolic regression (SR) recovers such models from
data~\citep{Cranmer2023pysr, LaCava2021srbench}.
Differentiable SR over a fixed tree of operator
nodes~\citep{Sahoo2018eql, Dong2024eeql, Odrzywolek2026eml} is
especially relevant when the learned expression is intended to
be inspected, simplified, or reused in downstream
signal-processing models.  The tree's structure encodes inductive
biases the practitioner controls before training.  The tree's
shape and where input variables enter the tree together form its
architecture, fixed before any data is seen.  How much that
fixed choice constrains gradient-based recovery, when operator
and grammar are held constant, remains difficult to isolate
because existing comparisons usually vary architecture together
with at least one other factor.

% Par 2: Related work, compressed
SRBench~\citep{LaCava2021srbench} compares 14 methods on 252
problems whose operator sets and search algorithms vary across
methods, confounding architecture with grammar and search.  The
Equation Learner (EQL)~\citep{Sahoo2018eql} and its extension
eEQL~\citep{Dong2024eeql} each evaluate a single fixed
architecture, leaving the effect of architecture choice
unmeasured.  MetaSymNet~\citep{Li2025metasymnet} and
DySymNet~\citep{Li2024dysymnet} adapt architecture during search,
quantifying end-to-end recovery rather than the contribution of
any one fixed architecture.  Methods outside the fixed-tree
gradient setting~\citep{Petersen2021dsr, Cranmer2023pysr,
Brunton2016sindy} operate on a different design space and are
not directly comparable.

% Par 3: Approach
Isolating variable-routing architecture requires a controlled
operator family in which the tested targets are representable
within the chosen depth limit.  The Exp-Minus-Log (EML) operator~\citep{Odrzywolek2026eml}
provides such a controlled setting for the targets studied here.  Three architectures, each a different
variable-routing pattern over a depth-3 EML tree, are compared
on a common set of targets, and two further operators, one
sharing EML's gradient asymmetry and one reversing that
asymmetry, test whether the findings depend on EML specifically
or on the gradient-asymmetry pattern exhibited by the operator.
This paper makes three contributions.  First, we formalize
recoverability as distinct from representability in fixed-tree
differentiable SR.  Second, we introduce a controlled comparison
in which variable-routing architecture is varied while operator
and grammar are held fixed.  Third, we show, as a proof of
concept, that held-out validation over a small architecture set
can mitigate some recovery failures, including on a Shockley
diode target outside the synthetic target family.
