%% ===================================================================
\section{Methodology}
\label{sec:setup}
%% ===================================================================

To test whether the architecture--target interaction is specific to
EML or general to operators with gradient asymmetry, two additional
operators are introduced:
\begin{align}
  \sml(a,b) &= \sinh(a) - \arctan(b), \label{eq:sml} \\
  \rml(a,b) &= \arctan(a) - \sinh(b). \label{eq:rml}
\end{align}
SML shares EML's gradient direction.  The left input receives
unbounded gradients ($\partial\sml/\partial a = \cosh a$, which
grows exponentially) while the right is bounded
($\partial\sml/\partial b = -1/(1+b^2)$).  RML reverses this
asymmetry.  Neither SML nor RML is known to be universal, and target trees are evaluated numerically.

Each node's left and right inputs are weighted combinations of
an architecture-specific set of available terms, typically drawn from
$\{1, x, y, f_{\text{child}}\}$.  Which terms are available at which
nodes is exactly what distinguishes the three architectures
(Table~\ref{tab:archs}).  Weights are controlled by learnable
parameters under a softmax or sigmoid.  Training minimizes mean
squared error using Adam with a learning rate
of 0.01 on a $21 \times 21$ grid over the positive quadrant
$[1,3]^2$ (441~points).  The domain is bounded away from zero
for two reasons.  First, EML's right-input logarithm $\ln b$
requires $b>0$.  Second, the singularity in
$\partial\eml/\partial b = -1/b$ near zero would destabilize
gradient measurements.  Targets with
numerical overflow on a subset of the grid are excluded entirely,
leaving only targets whose function values are finite on all
441~points.  Training proceeds in two phases.  A
search phase runs for 6{,}000 iterations at temperature $\tau = 2.5$, with
soft softmax and sigmoid weights, followed by a hardening phase
of 2{,}000 iterations in which $\tau$ anneals from 2.5 to 0.01 with entropy
and binarity penalties before the one-hot snap.  After hardening
each node selects exactly one input, producing a concrete
expression tree.

Each combination of architecture, operator, tree shape, and leaf
assignment is one cell.  V16 and Hybrid are tested with 256
trials per cell, using 64 seeds over four initialization
strategies.  Eq.\,6 is tested with 64 trials per cell using the
original paper's randn~$\times$~0.1 initialization.  Applying
V16's initialization set to Eq.\,6 would mix architecture and
initialization effects.  V16 and Hybrid initializations bias the
softmax weights either toward the constant~$1$ (\emph{biased}),
uniformly across inputs (\emph{uniform}), toward the variable
inputs (\emph{$xy$-biased}), or onto a single randomly chosen
input (\emph{random-hot}).  All reported rates carry 95\% Clopper--Pearson
confidence intervals (at $n = 256$,
mid-range half-widths $\approx$6\,pp; at $n = 64$,
0\% yields $[0, 5.6]$\%).  A trial counts as exact recovery when
the hardened model satisfies training MSE below $10^{-20}$ on the
441-point grid and held-out RMSE below $10^{-6}$ on 500 points
drawn uniformly from $[0.5,5.0]^2$, which strictly contains the
training grid.  In practice, this tolerance separates exact
symbolic recovery from close numerical fits on the noiseless
grid, and the held-out grid checks both in-range fit and
extrapolation.
Per-trial logs (twelve fields including seed, initialization,
architecture, target, operator, training and held-out RMSE,
hardened expression, routing, structural-match flag, failure-type
label, and wall time) are recorded for every run.

A sensitivity screen varied six protocol factors: optimizer,
hardening schedule, clamp magnitude, held-out grid seed,
training noise $\sigma$, and numerical $\varepsilon$.  The screen
ran 5{,}248 OFAT trials plus a factorial across the twelve
non-balanced chain cells.  The largest cell-dependent effects
came from initialization strategy.  One harmful level appeared
for the optimizer, Adam~$10^{-3}$, and one appeared for the
hardening schedule, step annealing.  The remaining levels
changed recovery by less than $12$ percentage points.  The main experiment
uses Adam~$10^{-2}$, power-law hardening, clamp~$=10^6$, grid
seed~$=12345$, $\sigma=0$, and $\varepsilon=10^{-12}$.\footnote{Source
code: \url{https://github.com/ChakshuGupta13/lab/}.}
